---
output: html_document
params:
  year: '2020'
---

# `r params$year` Encounter Report for Rhode Island Hospital 

Encounter reasons and visits by date from a simulated population of 10k Rhode Islanders. 

Data simulated using synthea: https://github.com/synthetichealth/synthea

```{r include=FALSE}
library(tidyverse)
library(lubridate)

year <- params$year
hospital <- 'rhode island hospital'

data <- read_csv('~/Foundation-of-the-R-Workflow-workshop-2019-09-06/Examples/data/simulated_patient_data.csv') %>% 
  mutate(start_date = as_date(start)) %>% 
  filter(year(start)==year, name == hospital)

visits_by_date <- data %>% 
  count(name,start_date)

visits_by_reason <- data %>% 
           count(reasondescription)

top_condition <- data %>% count(reasondescription) %>% drop_na() %>%  top_n(1) %>% pull(reasondescription)

top_encounter_type <- data %>% count(description) %>% drop_na() %>%  top_n(1) %>% pull(description)
```

## Overview
*  **`r n_distinct(data$patient)`** individuals visiting
* **`r top_encounter_type`** was the top encounter type
*   **`r top_condition`** was the top condition targeted during encounters  

## Top 10 reasons for visit

```{r echo=FALSE}
visits_by_reason %>% top_n(10, n) %>% drop_na() %>% 
  ggplot()+
  geom_col(aes(y=reorder(reasondescription, n), x=n), fill='mediumblue')+
  theme_classic()+
  xlab('Number of Encounters')+
  ylab('')
```

## Encounters by date

```{r echo=FALSE}
ggplot(visits_by_date)+
  geom_point(aes(x=start_date, y=n))+
  xlab('Date')+
  ylab('Number of Encounters')+
  theme_classic()
```
  
## Top Encounter Types  

```{r echo=FALSE}
data %>% count(description) %>% drop_na() %>%  top_n(10,n) %>% 
  ggplot()+
  geom_col(aes(y=reorder(description, n), x=n), fill='darkgreen')+
  theme_classic()+
  xlab('Number of Encounters')+
  ylab('')
```

