---
title: "function_tutorial_complete"
output: html_document
---

# Part 1
Creating our first R function. Hello World!
```{r}

hello_world<-function(){
  print('Hello, World')
}

```

Now let's execute our first function 
```{r}
hello_world()
```

Let's experiment with adding arguments. 
```{r}
hello_you <- function(name='You') {
  paste('Hello', name)
}

hello_you()
hello_you('Shannon')
```



# Part 2 Construct a function for our common 
This next block of code will load in the data sets we're going to use.
```{r}
library(tidyverse)
temp_df<-read_csv('temp_data.csv')
additive_df<-read_csv('additive_data.csv')

```

Below is the Code to convert a df to a summary df. The variabile we want to adjust for different data sets is `df`
```{r}
time <- max(df$Time)
df %>% 
  pivot_wider(names_from = Time, values_from = Value) %>% 
  rename(Time_0='0', Time_24='24') %>%
  mutate(Rate= Time_24-Time_0/time) %>% 
  group_by(Treatment) %>% 
  summarize(Average_rate = mean(Rate), Std_Dev=sd(Rate), n=n())
```


The skeleton of a function, we can get this by typing `fun` into the code block and hitting tab. Now you want to paste in the code we used to create the summary. You'll want to change the name and variable to fit the code.
```{r}
my_function <- function(df) {
  time <- max(df$Time)
df %>% 
  pivot_wider(names_from = Time, values_from = Value) %>% 
  rename(Time_0='0', Time_24='24') %>%
  mutate(Rate= Time_24-Time_0/time) %>% 
  group_by(Treatment) %>% 
  summarize(Average_rate = mean(Rate), Std_Dev=sd(Rate), n=n())
}
```

Try applying the function to the temp_df to create temp_df_summary
```{r}
my_function(temp_df)
```

Now let's apply the function to get summary data set for the additive experiment data
```{r}
my_function(additive_df)
```

