---
title: "Writing Functions In R"
output: html_document
---


# Part 1
Creating our first R function. Hello World!
```{r}

hello_world<-function(){
  print('Hello, World')
}

```

Now let's execute our first function 
```{r}
hello_world()
```

Let's experiment with adding arguments. 
```{r}

paste('Hello', name) # this is the action we want done for different names

```



# Part 2
Set up read in df
```{r}
temp_df<-read_csv('temp_data.csv')
additive_df<-read_csv('additive_data.csv')

```

# Here is the Code to convert the temp_df to our summary df
```{r}
time <- max(f$Time)
df %>% 
  pivot_wider(names_from = Time, values_from = Value) %>% 
  rename(Time_0='0', Time_24='24') %>%
  mutate(Rate= Time_24-Time_0/time) %>% 
  group_by(Treatment) %>% 
  summarize(Average_rate = mean(Rate), Std_Dev=sd(Rate), n=n())
```


The skeleton of a function
```{r}
name <- function(variables) {
  
}
```

Try applying the function to the temp_df to create temp_df_summary
```{r}
# Execute function with temp_df as argument variable
```

Now let's apply the function to get summary data set for the additive experiment data
```{r}
# Apply the function
```

