---
title: "Visualize Data"
---

## Your Turn 0

Add a setup chunk that loads the tidyverse packages.

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.



```{r}
mpg
```


## Your Turn 1

Run the code below to make a graph. Pay strict attention to spelling, capitalization, and parentheses!

```{r}
ggplot(data = mpg) +
 geom_point(mapping = aes(x = displ, y = hwy))
```

## Your Turn 2

Add `color`, `size`, `alpha`, and `shape` aesthetics to your graph. Experiment.  

size = class
shape = class
alpha = class
color = class

```{r}

```


## Your Turn 3

Replace this scatterplot with one that draws boxplots. Use the cheatsheet. Try your best guess.



```{r}

```

## Your Turn 4

Make a histogram of the `hwy` variable from `mpg`. Hint: do not supply a y variable.

```{r}

```

## Your Turn 5

Use the help page for `geom_histogram` to make the bins 2 units wide.

```{r}

```

## Your Turn 6

Make a bar chart `hwy` colored by `class`. Use the help page for `geom_bar` to choose a "color" aesthetic for class.

```{r}

```

## Quiz

What will this code do?

ggplot(mpg) + 
  geom_point(aes(displ, hwy)) +
  geom_smooth(aes(displ, hwy))

```{r}

```

## Quiz

What is different about this plot? Run the code!

```{r}
p <- ggplot(mpg) + 
  geom_point(aes(displ, hwy)) +
  geom_smooth(aes(displ, hwy))

library(plotly)
ggplotly(p)
```

## Your Turn 7

What does `getwd()` return?

```{r}

```

***

# Take aways

You can use this code template to make thousands of graphs with **ggplot2**.

```{r eval = FALSE}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```